% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
% !TeX program = pdfLaTeX
\documentclass[12pt]{article}
\usepackage{amsmath}
\usepackage{graphicx,psfrag,epsf}
\usepackage{enumerate}
\usepackage[]{natbib}
\usepackage{textcomp}


%\pdfminorversion=4
% NOTE: To produce blinded version, replace "0" with "1" below.
\newcommand{\blind}{0}

% DON'T change margins - should be 1 inch all around.
\addtolength{\oddsidemargin}{-.5in}%
\addtolength{\evensidemargin}{-1in}%
\addtolength{\textwidth}{1in}%
\addtolength{\textheight}{1.7in}%
\addtolength{\topmargin}{-1in}%

%% load any required packages here



% tightlist command for lists without linebreak
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}




\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\hypersetup{
  pdftitle={Maximizing Fairness with Synthetic data: Access to Emergency Fund in Sub-Saharan Africa},
  pdfkeywords={Fairness; predictive machine learning; emergency funds;
financial inclusion; ethics; gender bias; debiasing; synthetic data;
Sub-Saharan Africa; fairness metrics; Global Financial Inclusion
Database},
  hidelinks,
  pdfcreator={LaTeX via pandoc}}



\begin{document}


\def\spacingset#1{\renewcommand{\baselinestretch}%
{#1}\small\normalsize} \spacingset{1}


%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\if0\blind
{
  \title{\bf Maximizing Fairness with Synthetic data: Access to
Emergency Fund in Sub-Saharan Africa}

  \author{
        Charavee Basnet Chettri \\
    \\
     and \\     Vivian Wei \\
    \\
     and \\     Betty Pu \\
    \\
     and \\     Ziyue Yang \thanks{We are grateful to the Women at the
Table team and Sofia Kypraiou for the project inspiration and suggesting
the path forward} \\
    Department of Statistical and Data Sciences, Smith College,
Northampton MA\\
      }
  \maketitle
} \fi

\if1\blind
{
  \bigskip
  \bigskip
  \bigskip
  \begin{center}
    {\LARGE\bf Maximizing Fairness with Synthetic data: Access to
Emergency Fund in Sub-Saharan Africa}
  \end{center}
  \medskip
} \fi

\bigskip
\begin{abstract}
Financial inclusion is paramount for economic stability and resilience,
particularly in diverse regions like Sub-Saharan Africa, spanning low to
high-income countries and encompassing both resource-intensive and
non-resource-intensive economies. This study focuses on a crucial aspect
of financial resilience: the accessibility of emergency funds, defined
as having access to 1/20 of Gross National Income (GNI) per capita in
local currency within 30 days. Leveraging previous colleagues'
exploratory work on the Global Financial Inclusion Database 2021, our
objective is to mitigate inherent gender biases in the dataset by
rebalancing it with synthetic data, thereby enhancing fairness in
predicting emergency fund accessibility. Through predictive machine
learning modeling, we aim to contribute to the economic empowerment of
individuals in Sub-Saharan Africa, ultimately fostering resilience and
reducing disparities in access to essential financial resources.
\end{abstract}

\noindent%
{\it Keywords:} Fairness; predictive machine learning; emergency funds;
financial inclusion; ethics; gender bias; debiasing; synthetic data;
Sub-Saharan Africa; fairness metrics; Global Financial Inclusion
Database

\vfill

\newpage
\spacingset{1.9} % DON'T change the spacing!

\hypertarget{introduction}{%
\section{Introduction}\label{introduction}}

In the realm of financial inclusion, the accessibility of emergency
funds plays a pivotal role in determining an individual's financial
stability and resilience, especially in developing
countries.\citep{10.1093/oso/9780198827535.003.0007} In this project,
our goal is to predict the possibility for people in Sub-Saharan African
countries to come up with emergency funds, defined as 1/20 of GNI per
capita in local currency, within a 30-day
period\citep{Demirguc-Kunt2022}. This prediction serves as a crucial
factor for establishing future public financial policies, such as
determining the eligibility of individuals for loans and financial
assistance. The significance of this problem lies within its direct
impact on the economic well-being and empowerment of individuals in
developing regions. According to the
\href{https://www.worldbank.org/en/publication/globalfindex}{Global
Financial Inclusion (Global Findex) Database 2021} published by the
World Bank, only a little over half of people over 15 years of age in
developing economies could access extra funds within 30 days if faced
with unexpected expenses \citep{Demirguc-Kunt2022}. Therefore, there is
a pressing need to understand the factors influencing this accessibility
and eliminate inherent bias in the dataset. By delving into this issue,
we not only contribute to enhancing financial inclusion but also aid in
mitigating the different effects of financial shocks on vulnerable
populations.

Defining fairness is essential since the concept itself is relative
among different people. In the data science discourse, fairness
encompasses three key aspects: individual fairness, group fairness, and
causal fairness\citep{Kypraiou2021What}. Individual fairness focuses on
preventing discrimination against individuals with similar relevant
characteristics. This means ensuring that individuals in similar
situations receive similar outcomes from the model, regardless of
irrelevant factors\citep{10.1145/3461702.3462621}. Group fairness aims
to prevent disparities in outcomes for different groups. This ensures
equal opportunities for all groups, regardless of their
membership\citep{10.1145/3442188.3445876}. Causal fairness goes beyond
simply observing disparities and delves into understanding their
underlying causes. It seeks to mitigate these root causes to achieve
fair outcomes within and across groups\citep{plecko2022causal}.

Our previous colleagues conducted analysis using the \textless AI \&
Equality\textgreater{} Human Rights Toolbox and used a Decision Tree
Classifier machine learning model implemented via Python to predict
access to emergency funds with 68\% accuracy. Their work laid a solid
foundation by exploring demographic and financial variables within the
dataset, and assessed the fairness of the decision tree classifier,
particularly concerning gender bias, and then applied various processing
techniques to enhance the fairness of the model\citep{Porta2022}.

Based on their work, our group aims to incorporate synthetic data to
rebalance the dataset, ensuring equitable representations amongst both
genders in the dataset. Our approach aims to consider a broader
selection of machine learning algorithms and mitigate the disparities
the previous colleagues found in the decision tree's classifier's
predictions and enhance fairness with synthetic data, ultimately better
predict access to emergency funds in south-Saharan Africa
countries\citep{SyntheticDataAIEquality}.

\hypertarget{background-on-sub-saharan-region}{%
\section{Background on Sub-Saharan
Region}\label{background-on-sub-saharan-region}}

Sub-Saharan Africa is a region characterized by a diverse economic
landscape, encompassing low, lower-middle, and upper-middle-income
countries. Demographically, Sub-Saharan Africa is marked by a rich
tapestry of cultures and a population exceeding 1.2 billion people. This
expansive and diverse demographic landscape includes 22 countries
grappling with fragility or conflict, posing unique challenges to
development efforts. Additionally, 13 small states within the region are
characterized by limited human capital, modest populations, and
constrained land areas.

According to the World Bank's definition, middle-income countries had a
per capita gross national income of more than US\$995.00 in the years
2015--17. Among the 35 countries included in our dataset, 20 are
classified as low-income countries, and 15 are classified as
middle-income countries. Additionally, 11 are classified as countries in
Fragile and Conflict-Affected Situations, which, by definition, have
experienced a peacekeeping or peace-building mission within the last
three years.

Sub-Saharan African countries not only differ in terms of economic
prosperity but also in economic structure and resource intensity.
Resource-intensive countries include both oil-exporting nations, where
net oil exports make up 30 percent or more of total exports, and
commodity exporters, where nonrenewable natural resources represent 25
percent or more of total exports. The divergence between
resource-intensive and non-resource-intensive countries became more
entrenched following the commodity price shock of 2015.
Non-resource-intensive countries have proven more resilient, supported
by their more diversified economies. On the other hand,
resource-intensive economies generally have a less diversified
structure, making them more susceptible to external shocks. In our
dataset, 6 of the countries are oil exporters, and 11 export other
commodities such as iron ore, copper, cotton, coffee, and sugar. The
remaining 18 countries are non-resource-intensive and their economies
are not reliant on exports.

In recent years, the Sub-Saharan Africa region has grappled with
significant economic challenges, including soaring inflation, pronounced
exchange rate pressures, debt vulnerabilities, and widening economic
disparities within the region. Therefore, addressing these structural
and economic disparities is imperative for tackling developmental issues
in the region.

\hypertarget{data}{%
\section{Data}\label{data}}

In order to continue assessing and enhancing the fairness of the machine
learning models, we use the same Global Findex Database as our
colleagues before us. The Global Findex database {[} {]} was first
launched in 2011 by the World Bank---with funding from the Bill \&
Melinda Gates Foundation. It is the world's most comprehensive data set
on how adults save, borrow, make payments, and manage risk. The dataset
contains over 200 indicators including account ownership, payments,
savings, credit, and financial resilience and has coverage over 140
nations, representing 97\% of the world's population for year 2017,
2014, and 2011.

The data was constructed through a series of surveys carried out by
Gallup, Inc in association with the annual Gallup World Poll. They
randomly sampled 1000 individuals from each country and asked them to
respond to a survey either over the phone or in-person. The target
population is the civilian, non-institutionalized population 15 years
and above. In consistency with the former colleagues, our sample covers
only the Sub-Saharan region (35 countries total), thus there are 35,000
total observations and 105 total variables. Since the sampling was
random and the sample size is large, we can assume that our sample is
representative of the total population of people living in the 35
Sub-Saharan countries in the data. The data was collected directly from
individuals over the 2017 calendar year and is self-contained, meaning
it does not rely on any external resources.

\hypertarget{variable-definition-and-descriptions}{%
\subsection{Variable Definition and
Descriptions}\label{variable-definition-and-descriptions}}

After basic data cleaning, we have created various visualizations to
help understand who is in the data and acknowledge potential sources of
bias within the data.

\emph{Outcome Variable of Interest: Access to Emergency Fund (Fin24)}

The variable ``Fin24'' in the dataset asked participants the question:
Now, imagine that you have an emergency and you need to pay {[}1/20 of
GNI per capita in local currency{]}. Is it possible or not possible that
you could come up with {[}1/20 of GNI per capita in local currency{]}
within the NEXT MONTH? In order to make it more straightforward for
interpretations, we restrict the response into a binary variable for Yes
and No.~The overall distribution of access to emergency funds showed 17,
599 individuals had access to emergency funds while 14342 did not. This
indicates that over half of individuals represented in the data do not
have access to emergency funds.

To help conceptually understand how the outcome variables might vary
among different levels of the variables, we chose 50\% as the benchmark
proportion for checking if the possibility of coming up with emergency
funds for each variable group of interest is different from a random
50/50 chance.

\bibliographystyle{plain}
\bibliography{bibliography.bib}



\end{document}
